dbRDA analysis

### 2017-08-02 doc/RMD/RDA-analysis.Rmd
update RDA to include MCG

# function to gather Environmental variables, but averaging Severity
# (aka Straw Test, Virulence). This function is necessary because the
# data tends to randomly shuffle when being processed vai dplyr functions
# for better or for worse. ¯\_(ツ)_/¯
# 
# @param DAT the full data set
# @param CC The clone-corrected data set
# @param wmn should the "Source" column be converted to a binary wmn
#   factor? When this is true, everything that is not wmn will be converted
#   to "other"
# 
# @return A data frame containing Severity, Host, Source, Year, Region, and MLG
makeENV <- function(DAT, CC, wmn = FALSE){
  # Creating the data frame with severity
  META   <- select(other(DAT)$meta, -Isolate)
  STRATA <- select(strata(DAT), -MCG)
  STRATA <- if (wmn) mutate(STRATA, Source = ifelse(Source == "wmn", "wmn", "other")) else STRATA
  MLL    <- data.frame(MLG = mll(DAT))
  sev    <- bind_cols(META, STRATA, MLL) %>%
    group_by(Host, Source, Year, Region, MLG) %>% 
    summarize(Severity = mean(Severity)) %>% # Get mean severity per MLG
    ungroup()
  # Ensuring the data is in the correct order
  META   <- select(other(CC)$meta, -Isolate)
  STRATA <- select(strata(CC), -MCG)
  MLL    <- data.frame(MLG = mll(CC))
  bind_cols(META, STRATA, MLL) %>% 
    left_join(sev)
}

@@

# This was ganked from https://github.com/gavinsimpson/ggvegan/blob/59d233977a5b2d15d4de150b782fb1794aa1de8b/R/utils.R
# 
# @title Scale Vectors to Data
# @description Scale vector arrows to \code{fill} proportion of the data.
# @param arrows a two-column matrix-like object containing coordinates for the arrows/vectors on x and y axes.
# @param data a two-column matrix-like object containing coordinates of the data on the x and y axes.
# @param at numeric vector of length 2; location of the origin of the arrows.
# @param fill numeric; what proportion of the range of the data to fill
# @return a numeric multiplier that will scale the arrows
# @author Gavin L. Simpson
arrowMul <- function(arrows, data, at = c(0, 0), fill = 0.75) {
    u <- c(range(data[,1], range(data[,2])))
    u <- u - rep(at, each = 2)
    r <- c(range(arrows[, 1], na.rm = TRUE), range(arrows[, 2], na.rm = TRUE))
    rev <- sign(diff(u))[-2]
    if (rev[1] < 0)
        u[1:2] <- u[2:1]
    if (rev[2] < 0)
        u[3:4] <- u[4:3]
    u <- u/r
    u <- u[is.finite(u) & u > 0]
    fill * min(u)
}
# Plotting the dbRDA results
# 
# @param db a capscale object
# @return a ggplot2 object from the scores 
plot_dbrda <- function(db, arrows = 10, seed = 2017-06-28){
  set.seed(seed)
  dbsum     <- scores(db, display = c("cn", "bp", "sites"), scaling = "sites")
  Centroids <- as.data.frame(dbsum$centroids)
  Centroids <- rownames_to_column(Centroids, var = "cent_type")
  Centroids <- mutate_(Centroids, .dots = list(Length = ~sqrt(CAP1^2 * CAP2^2)))
  # Centroids
  SampleCentroids <- rownames_to_column(data.frame(dbsum$sites), var = "isolate_names")
  labs    <- summary(db, axes = 0)[["cont"]][["importance"]]["Proportion Explained", 1:2]
  terms   <- paste0("(", paste(attr(db$terms, "term.labels"), collapse = "|"), ")")
  mul     <- arrowMul(dbsum$biplot[, 1:2], dbsum$sites)
  Arrows  <- data.frame(dbsum$biplot * mul)
  Arrows  <- rownames_to_column(Arrows, var = "class")
  # Making the classes presentable
  Arrows$class <- gsub(terms, "\\1: ", Arrows$class)
  Arrows$class <- gsub(": $", "", Arrows$class)
  Arrows$class <- gsub("unk", "unknown", Arrows$class)
  Arrows  <- mutate_(Arrows, .dots = list(Length = ~sqrt(CAP1^2 * CAP2^2)))
  Arrows  <- arrange(Arrows, Length)
  Arrows  <- top_n(Arrows, arrows)
  ggplot(Centroids, aes(x = CAP1, y = CAP2))+
    geom_point(data = SampleCentroids, 
               # alpha = 1/2, 
               # fill = "white",
               # fill = "dark orange", 
               color = "grey45", 
               size = 2.5, 
               pch = 21)+
    coord_cartesian() +
    geom_segment(aes(x = 0, xend = CAP1, 
                     y = 0, yend = CAP2),
                 arrow = arrow(length = unit(0.3, "cm")), 
                 data = Arrows
                 ) + 
    geom_label_repel(aes(x = CAP1, y = CAP2, label = class), 
                     point.padding = unit(0.5, "lines"),
                     segment.color = "grey25",
                     data = Arrows) +
    xlab(paste0("Eig 1 (", round(labs[[1]]*100, 2), "% variance explained)")) +
    ylab(paste0("Eig 2 (", round(labs[[2]]*100, 2), "% variance explained)"))
}
```

@@

```{r resultplot, fig.width = 5, fig.height = 5}
set.seed(999)
plot_dbrda(cap11cc, arrows = 5) + 
  theme_classic(base_size = 16, base_family = "Helvetica") + 
  theme(axis.text = element_text(color = "black")) +
  theme(aspect.ratio = 1)
FILE <- file.path(PROJHOME, "results", "figures", "publication", "Figure7Z.pdf")
ggsave(filename = FILE, width = 88, height = 88, units = "mm", scale = 1.5)
plot_dbrda(cap16cc, arrows = 10) +
  theme_classic(base_size = 16, base_family = "Helvetica") + 
  theme(axis.text = element_text(color = "black")) +
  theme(aspect.ratio = 1)
```

@@

```{r, echo = FALSE, results = "asis"}
cap11sum <- summary(cap11cc, axes = 0, scaling = "sites")
cap16sum <- summary(cap16cc, axes = 0, scaling = "sites")
vartab <- function(x){
  tot <- x$tot.chi
  con <- x$constr.chi
  uco <- x$unconst.chi
  data.frame(list(Inertia = c(tot, con, uco),
                  Proportion = c(tot, con, uco)/tot))
}
cat("\n\n11 loci:\n\n")
knitr::kable(vartab(cap11sum), digits = 3)
cat("\n\n16 loci:\n\n")
knitr::kable(vartab(cap16sum), digits = 3)
```

@@

```{r}
dat11raw <- genind2df(dat11cc, usepop = FALSE) %>% mutate_all(funs(as.integer(.)))
dat16raw <- genind2df(dat16cc, usepop = FALSE) %>% mutate_all(funs(as.integer(.)))
vp11     <- varpart(dat11cc.bruvo, ~Severity + Year + Region + Host, data = ENV11, comm = dat11raw, add = TRUE)
plot(vp11, Xnames = c("Full Model", "No Model"))
try(vp16     <- varpart(dat16cc.bruvo, ~Year + Region + Host, data = ENV16, comm = dat16raw, add = TRUE))
```

```{r}
vp11 <- varpart(dat11cc.bruvo, ~Severity, ~Year, ~Region, ~Host, data = ENV11, add = TRUE)
vp16 <- varpart(dat16cc.bruvo, ~Year, ~Region, ~Host, data = ENV16, add = TRUE)

plot(vp11, digits = 2, Xnames = c("Severity", "Year", "Region", "Host"))

plot(vp16, digits = 2, Xnames = c("Year", "Region", "Host"))
```

>>>>

# function to gather Environmental variables, but averaging Severity
# (aka Straw Test, Virulence). This function is necessary because the
# data tends to randomly shuffle when being processed vai dplyr functions
# for better or for worse. ¯\_(ツ)_/¯
# 
# @param DAT the full data set
# @param CC The clone-corrected data set
# @param wmn should the "Source" column be converted to a binary wmn
#   factor? When this is true, everything that is not wmn will be converted
#   to "other"
# 
# @return A data frame containing Severity, Host, Source, Year, Region, and MLG
makeENV <- function(DAT, CC, wmn = FALSE){
  # Creating the data frame with severity
  META   <- select(other(DAT)$meta, -Isolate)
  STRATA <- strata(DAT)
  STRATA <- if (wmn) mutate(STRATA, Source = ifelse(Source == "wmn", "wmn", "other")) else STRATA
  MLL    <- data.frame(MLG = mll(DAT))
  sev    <- bind_cols(META, STRATA, MLL) %>%
    group_by(Host, Source, Year, Region, MLG) %>% 
    summarize(Severity = mean(Severity)) %>% # Get mean severity per MLG
    ungroup()
  # Ensuring the data is in the correct order
  META   <- select(other(CC)$meta, -Isolate)
  STRATA <- strata(CC)
  MLL    <- data.frame(MLG = mll(CC))
  bind_cols(META, STRATA, MLL) %>% 
    left_join(sev)
}

@@

# This was ganked from https://github.com/gavinsimpson/ggvegan/blob/59d233977a5b2d15d4de150b782fb1794aa1de8b/R/utils.R
# 
# @title Scale Vectors to Data
# @description Scale vector arrows to \code{fill} proportion of the data.
# @param arrows a two-column matrix-like object containing coordinates for the arrows/vectors on x and y axes.
# @param data a two-column matrix-like object containing coordinates of the data on the x and y axes.
# @param at numeric vector of length 2; location of the origin of the arrows.
# @param fill numeric; what proportion of the range of the data to fill
# @return a numeric multiplier that will scale the arrows
# @author Gavin L. Simpson
arrowMul <- function(arrows, data, at = c(0, 0), fill = 0.75) {
    u <- c(range(data[,1], range(data[,2])))
    u <- u - rep(at, each = 2)
    r <- c(range(arrows[, 1], na.rm = TRUE), range(arrows[, 2], na.rm = TRUE))
    rev <- sign(diff(u))[-2]
    if (rev[1] < 0)
        u[1:2] <- u[2:1]
    if (rev[2] < 0)
        u[3:4] <- u[4:3]
    u <- u/r
    u <- u[is.finite(u) & u > 0]
    fill * min(u)
}
# Plotting the dbRDA results
# 
# @param db a capscale object
# @return a ggplot2 object from the scores 
plot_dbrda <- function(db, arrows = 10, seed = 2017-06-28, lab = TRUE){
  set.seed(seed)
  dbsum     <- scores(db, display = c("cn", "bp", "sites"), scaling = "sites")
  Centroids <- as.data.frame(dbsum$centroids)
  Centroids <- rownames_to_column(Centroids, var = "cent_type")
  Centroids <- mutate_(Centroids, .dots = list(Length = ~sqrt(CAP1^2 * CAP2^2)))
  # Centroids
  SampleCentroids <- rownames_to_column(data.frame(dbsum$sites), var = "isolate_names")
  if (lab){
    labs    <- vegan:::summary.cca(db, axes = 0)[["cont"]][["importance"]]["Proportion Explained", 1:2]
    xl      <- paste0("Eig 1 (", round(labs[[1]]*100, 2), "% variance explained)")
    yl      <- paste0("Eig 2 (", round(labs[[2]]*100, 2), "% variance explained)")
  } else {
    xl      <- "Eig 1"
    yl      <- "Eig 2"
  }
  terms   <- paste0("(", paste(attr(db$terms, "term.labels"), collapse = "|"), ")")
  mul     <- arrowMul(dbsum$biplot[, 1:2], dbsum$sites)
  Arrows  <- data.frame(dbsum$biplot * mul)
  Arrows  <- rownames_to_column(Arrows, var = "class")
  # Making the classes presentable
  Arrows$class <- gsub(terms, "\\1: ", Arrows$class)
  Arrows$class <- gsub(": $", "", Arrows$class)
  Arrows$class <- gsub("unk", "unknown", Arrows$class)
  Arrows  <- mutate_(Arrows, .dots = list(Length = ~sqrt(CAP1^2 * CAP2^2)))
  Arrows  <- arrange(Arrows, Length)
  Arrows  <- top_n(Arrows, arrows)
  ggplot(Centroids, aes(x = CAP1, y = CAP2))+
    geom_point(data = SampleCentroids, 
               # alpha = 1/2, 
               # fill = "white",
               # fill = "dark orange", 
               color = "grey45", 
               size = 2.5, 
               pch = 21)+
    coord_cartesian() +
    geom_segment(aes(x = 0, xend = CAP1, 
                     y = 0, yend = CAP2),
                 arrow = arrow(length = unit(0.3, "cm")), 
                 data = Arrows
                 ) + 
    geom_label_repel(aes(x = CAP1, y = CAP2, label = class), 
                     point.padding = unit(0.5, "lines"),
                     segment.color = "grey25",
                     data = Arrows) +
    xlab(xl) +
    ylab(yl)
}
```

@@

```{r resultplot, fig.width = 5, fig.height = 5}
set.seed(999)
plot_dbrda(cap11cc, arrows = 5) + 
  theme_classic(base_size = 16, base_family = "Helvetica") + 
  theme(axis.text = element_text(color = "black")) +
  theme(aspect.ratio = 1)
# if (!interactive()){
  FILE <- file.path(PROJHOME, "results", "figures", "publication", "Figure7Z.pdf")
  ggsave(filename = FILE, width = 88, height = 88, units = "mm", scale = 1.5)
# }
plot_dbrda(cap16cc, arrows = 10) +
  theme_classic(base_size = 16, base_family = "Helvetica") + 
  theme(axis.text = element_text(color = "black")) +
  theme(aspect.ratio = 1)
```

@@

```{r, echo = FALSE, results = "asis"}
cap11sum <- vegan:::summary.cca(cap11cc, axes = 0, scaling = "sites")
cap16sum <- vegan:::summary.cca(cap16cc, axes = 0, scaling = "sites")
vartab <- function(x){
  tot <- x$tot.chi
  con <- x$constr.chi
  uco <- x$unconst.chi
  data.frame(list(Inertia = c(tot, con, uco),
                  Proportion = c(tot, con, uco)/tot))
}
cat("\n\n11 loci:\n\n")
knitr::kable(vartab(cap11sum), digits = 3)
cat("\n\n16 loci:\n\n")
knitr::kable(vartab(cap16sum), digits = 3)
```

@@

```{r}
dat11raw <- genind2df(dat11cc, usepop = FALSE) %>% mutate_all(funs(as.integer(.)))
dat16raw <- genind2df(dat16cc, usepop = FALSE) %>% mutate_all(funs(as.integer(.)))
vp11     <- varpart(dat11cc.bruvo, ~Year + Region + Host + MCG, data = ENV11, comm = dat11raw, add = TRUE)
plot(vp11, Xnames = c("Full Model", "No Model"))
try(vp16     <- varpart(dat16cc.bruvo, ~Year + Region + Host, data = ENV16, comm = dat16raw, add = TRUE))
```

```{r}
vp11 <- varpart(dat11cc.bruvo, ~Year, ~Region, ~Host, ~MCG, data = ENV11, add = TRUE)
vp16 <- varpart(dat16cc.bruvo, ~Year, ~Region, ~Host, ~MCG, data = ENV16, add = TRUE)

plot(vp11, digits = 2, Xnames = c("Year", "Region", "Host", "MCG"))

plot(vp16, digits = 2, Xnames = c("Year", "Region", "Host", "MCG"))
```
